Genome-wide changes in expression during inflorescence architecture development

### 2015-10-02 ms.Rmd
Added discussion on clusters, GSEA and MADS.

knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               dev=c("png", "cairo_pdf"),
               fig.width=3.150,
               fig.height=8.661,
               fig.align = 'center',
               dpi=300,
               fig.show='hold',
               cache=FALSE,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

@@

| ^§^ Institut de Recherche pour le Développement (IRD), 911 Avenue Agropolis, 34394 Montpellier, France.
| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries. As a preliminary quality check, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`).

**Our in situ results could go here (or in SI).**

@@

Two G1-Like (ALOG) genes, *G1L2* and *G1L5* (*TAWAWA*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively. Only three ALOG genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). *G1L5* mutants **do what, incl. ref**, and the co-expression of other ALOG genes suggests redundancy within the family.

**Discussion: what might these patterns mean biologically? Which clusters have genes that are already candidates for panicle branching?**

@@

Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). Unlike soft clustering, which groups genes based on shared expression dynamics, this GSEA was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in expression of their members in each stage. These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development. The most pronounced change in expression appears to occur between the RM and branch meristem samples. Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types. Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.

The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).

**Discussion: In which samples were TF families involved in branching enriched?**

## Evolution of MADS genes in monocots and dicots

There was an overrepresentation of MADS genes in cluster `r getClusterName('small', 'inc')` and their overall expression is enriched in the spikelet meristem sample. These results correspond with previous studies that have established the importance of MADS genes in inflorescence development in rice and other plants (**expand**).

The change in expression of MIKC proteins between undifferentiated and differentiated meristems was extracted from the dataset reported here and two other studies that used LMD and RNA-sequencing in early stages of inflorescence development in Arabidopsis and tomato [@Mantegazza:2014di; @Park:2012dg]. To investigate the evolution of the MADS family in Angiosperms, these changes were compared to homology of MIKC proteins in the three species (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).

| **Figure**: Similarity tree of MADS proteins with L2FC values between comparable samples (and dN/dS values?) for each one:
| **rice**: rachis meristem vs. spikelet merisem.
| **arapidospis**: inflorescence meristem vs. floret meristem.
| **tomato**: sympodial IM vs. floret meristem.

@@

*Oryza sativa* ssp. *japonica* cv. Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night. After 7 weeks in long day conditions (LD; 14 hours light and 10 hours darkness) plants were transferred to short day conditions (SD; 10 hours light and 14 hours darkness) to induce floral transition. Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.

@@

Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (=**filtered**?) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C. Embedding for dissection was performed as described by @Mantegazza:2014di. Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 LASER dissector (Leica). RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN). Library preparation with **which kit** and 50-base single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).

>>>>

knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               dev=c("png", "cairo_pdf"),
               fig.width=3.150,
               fig.height=8.661,
               fig.align = 'center',
               dpi=300,
               fig.show='hold',
               cache=TRUE,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

@@

| ^§^ Institut de Recherche pour le Développement, 911 Avenue Agropolis, 34394 Montpellier, France.
| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries. As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**).

@@

The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes. Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively. Only three ALOG genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`). Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff]. *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff]. This might be explained by redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other ALOG genes, but I'm not sure how much is required for off-target effects.]

The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions. Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [**Komatsu 2003 FZP paper**], were found in clusters containing genes that are highly expressed in the SM.

| **What could any of these patterns mean biologically?**
| **I picked out some genes from the clusters (list directly below), are there any others in here I should discuss?**

@@

Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). Unlike soft clustering, which groups genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members. There are some notable modes of expression shared between different families that may indicate roles in inflorescence development. The most pronounced change in expression appears to occur between the RM and branch meristem samples. Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types. Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.

The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).

These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs, and mutants of the former gene produce fewer spikelets than wild-type plants [@Gao:2010iz] (**check role of PAP2**). In contrast to *PAP2*, *LHS1* was detected only in the SM samples, indicative of the overall pattern for other MADS genes.

## Comparative analysis of MADS gene expression in eudicots and monocots

The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by **ref: Prunet/Jack review**). In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**]. [**Add info about tomato MADS**].

* Something about MADS genes involved earlier in development?
* Expression of MADS genes that interact with *G1L5*? (Maybe this is too much detail?)

A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper. The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A. thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di]. A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`). [**Note about relative genome size (number of genes) and MADS family size between species**.]

As expected based on their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems. However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].

* *AGL6*-like: AGL6 has a defined function but it doesn't look like other genes in the clade
* *STMADS11*-like:
* *SQUA*-like:
* *DEF*-like:
* *GLO*-like:
* *FLC*-like:

@@

*Oryza sativa* ssp. *japonica* cv. Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night. After 7 weeks in long day conditions (14 hours light and 10 hours darkness) plants were transferred to short day conditions (10 hours light and 14 hours darkness) to induce floral transition. Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.

@@

Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (**filtered?**) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C. Embedding for dissection was performed as described by @Mantegazza:2014di. Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 LASER dissector (Leica). RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN). Library preparation with **which kit** and 50-base, single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
